---
Title: "Exploratory Data Analysis – Road‑Traffic Mortality (Simple Version)"
output: html_document
---

```{r setup, include=FALSE}
packages <- c("tidyverse", "naniar", "GGally", "skimr", "patchwork")

install.packages(setdiff(packages, rownames(installed.packages())))
lapply(packages, library, character.only = TRUE)

dat <- readr::read_csv("../data/clean/traffic_mortality_panel.csv",
                       show_col_types = FALSE)
```

## 1 Data overview

```{r skim}
skimr::skim(dat)
```

## 2 Missing‑data pattern

```{r missing}
# Bar chart of missing rate per variable
naniar::gg_miss_var(dat) +
  labs(title = "Missing rate by variable")

# UpSet plot – top 5 missing‑value sets
naniar::gg_miss_upset(dat, nsets = 5)
```

## 3 Subset for modelling

```{r subset}
# Keep observations from 2000 onwards where GDP & vehicles variables are available
# If you plan to drop vehicles_pc from the model, comment out the is.na(vehicles_pc) line.

dat_mod <- dat %>%
  dplyr::filter(year >= 2000,
                !is.na(gdp_pc),
                !is.na(vehicles_pc))
# Compare row counts
 tibble(version = c("All years", "Year >= 2000 & no NA veh/gdp"),
        rows    = c(nrow(dat), nrow(dat_mod)))
```

## 4 Univariate distributions

```{r dists, fig.height=5}
# Histograms (log‑scale where highly skewed)

p1 <- ggplot(dat, aes(deaths))      + geom_histogram(bins = 40) + scale_x_log10() + labs(title = "Deaths (log10)")
p2 <- ggplot(dat, aes(death_rate))  + geom_histogram(bins = 40) + labs(title = "Deaths per 100k")
p3 <- ggplot(dat, aes(gdp_pc))      + geom_histogram(bins = 40) + scale_x_log10() + labs(title = "GDP per cap (log10)")
p4 <- ggplot(dat, aes(vehicles_pc)) + geom_histogram(bins = 40) + labs(title = "Vehicles per 1k")

(p1 | p2) / (p3 | p4)
```

## 5 Time trend

```{r trend}
dat %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(median = median(death_rate, na.rm = TRUE),
                   q25    = quantile(death_rate, 0.25, na.rm = TRUE),
                   q75    = quantile(death_rate, 0.75, na.rm = TRUE)) %>%
  ggplot(aes(year, median)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = .2) +
  geom_line(size = 1) +
  labs(y = "Deaths per 100k", title = "Global trend: median & IQR")
```

## 6 Pairwise relationships (subset 2000+)

```{r pairs, fig.height=7, warning=FALSE}
if (nrow(dat_mod) > 30) {
  dat_mod %>%
    dplyr::select(death_rate, gdp_pc, vehicles_pc) %>%
    GGally::ggpairs(progress = FALSE)
}
```

## 7 Save 100‑row sample for the repository

```{r sample, echo=FALSE, results='hide'}
set.seed(447)
readr::write_csv(
  dplyr::sample_n(dat, size = min(nrow(dat), 100)),
  "../data/sample/traffic_sample.csv")
```

## 8 Takeaways

* Vehicles per‑capita is missing **`r round(mean(is.na(dat$vehicles_pc))*100,1)`%** of the time. Under **Model A**, we exclude this variable.
* Death rate shows a negative relationship with GDP per capita.
* We will compare:
  * **Frequentist baseline** – `glm.nb()` with `death ~ gdp_pc + offset(log(pop_total))`.
  * **Bayesian NB**  fitted in Stan.
  * **Metrics:** PSIS‑LOO for Bayesian, AIC & 10‑fold CV for frequentist.
```
